Compare permutation and scalar transformation approaches on simulated and real data

### 2021-03-03 doc/choosing_significant_harmonies.Rmd
cleaning folder

```{r, eval = FALSE, echo = FALSE}
set.seed(12345)

# simulate one big data which will have all categories generated on the #fly

@@

simtable<-read_csv(here::here('simulations/sim_table/sim_table.csv'))

lapply(1:2, 
                           function(scen){

simj<-simtable[scen,] 
nfacetj<-simj$nfacet
nxj<-simj$nx
nperm = 20

id_x <- rep_len(seq_len(nxj), length.out = lensim) 
id_facet <- rep_len(seq_len(nfacetj), length.out = lensim)

@@

wpd_orig = compute_pairwise_max(sim_panel_data, 
                                gran_x = "id_x",
                                gran_facet = "id_facet",
                                response = sim_data)

@@

  wpd =  compute_pairwise_max(sim_panel_sample, 
                              gran_x = "id_x",
                              gran_facet = "id_facet",
                              response = sim_data) %>% 
    as_tibble() %>% mutate(perm_id = x)

@@

wpd_all <- wpd_orig %>% as_tibble() %>% 
  mutate(perm_id = 0) %>% 
  bind_rows(wpd_sample)

saveRDS(wpd_all,
        paste0('simulations/supplementary/test/wpd_N01/data-ind/',
               nxj,'-',
               #"nfacetj-",
               nfacetj,"-",
               '-wpd-test.rds'))
})
```

The p-value of the design $D_{null}$ is size.
The p-value of other designs is power.
Confidence interval of the

>>>>

```{r, eval = FALSE, echo = FALSE}
sample_seed = seq(1000, 2999, by = 10)
nperm = 200
simtable<-read_csv(here::here('simulations/sim_table/sim_table.csv'))
# simulate one big data which will have all categories generated on the #fly

all_seed <- lapply(seq_len(sample_seed), function(x){

  set.seed(sample_seed[x])

@@

# for each row of the sim table
each_seed <- lapply(nrow(simtable), 
                           function(scen){

simj<-simtable[scen,] 
nfacetj<-simj$nfacet
nxj<-simj$nx

id_x <- rep_len(seq_len(nxj), length.out = lensim) 
id_facet <- rep_len(rep(seq_len(nfacetj), each = nxj), length.out = lensim)

@@

wpd_orig = compute_pairwise_norm_scalar(sim_panel_orig, 
                                gran_x = "id_x",
                                gran_facet = "id_facet",
                                response = sim_data)

@@

  wpd =  compute_pairwise_norm_scalar(sim_panel_sample, 
                              gran_x = "id_x",
                              gran_facet = "id_facet",
                              response = sim_data) %>% 
    as_tibble() %>% mutate(perm_id = x)

@@

wpd_all <- wpd_orig %>% as_tibble() %>% 
  mutate(perm_id = 0) %>% 
  bind_rows(wpd_sample) %>% 
  bind_cols(nx = nxj, nfacet = nfacetj)

wpd_all
}) %>% bind_rows()

# find threshold

threshold <- each_seed %>% mutate(
  threshold = quantile(each_seed$value, probs = 0.95))

# conduct test

threshold %>% 
  dplyr::filter(perm_id == 0) %>% 
  mutate(select = if_else(value>threshold, "yes", "no"))

}) %>% bind_rows(.id = "seed_id")

saveRDS(all_seed,
        paste0('simulations/supplementary/test/wpd_N01/null_all_seed_threshold.rds')

```

### 2021-03-03 doc/combining_normalisation_method.Rmd
cleaning folder

@@

G21_glm <- G21 %>% 
  mutate(wpd_glm =  (value - (1/(intercept + slope*log(nx*nfacet)
                    )
                    )
             ),
         wpd_glm_scaled = ((wpd_glm*300)))

>>>>

G21_sd  = G21 %>% 
    mutate(wpd_glm =  (value - (1/(intercept + slope*log(nx*nfacet)
    )
    )
    ))

scale_fac <- 1/G21_sd$wpd_glm %>% sd()

@@

G21_glm <- G21 %>% 
  mutate(wpd_glm =  (value - (1/(intercept + slope*log(nx*nfacet)
                    )
                    )
             ),
         wpd_glm_scaled = ((wpd_glm*320)))

G21_glm$wpd_glm_scaled %>% sd()

### 2021-03-03 doc/comparison_perm_scalar.Rmd
cleaning folder

---
title: "Permutation and scalar transformation approaches to normalisation and combining them"
author: "Sayani Gupta"
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      eval = TRUE,
                      warning = FALSE,
                      message = FALSE)
knitr::read_chunk("example_transformations.R")
library(tidyverse)
library(readr)
library(ggplot2)
library(here)
library(patchwork)
```

# Data generation

@@

Figure \ref{Fig:raw} shows the distribution of $wpd_{raw}$ is plotted across different nx and nfacet categories. Both shape and scale of the distribution changes for different $nx$ and $nfacet$ categories. This is not desirable as it would mean we would not be able to compare $wpd_{raw}$ across different $nx$ and $nfacet$ as each of them are drawn from distributions of different locations and scale. In Figure \ref{fig:quadratic} we see how the median of $wpd_{raw}$ varies with the total number of distances $nx_i*nfacet_j$ for a panel with $nx = nx_i$ categories and $nfacet = nfacet_j$ categories. The median increases abruptly for lower values of $nx_i*nfacet_j$ and the increase becomes more gradual for higher $nx_i*nfacet_j$.

@@

```{r quadratic, fig.cap = "$wpd_{raw}$ is plotted against nx*nfacet and the blue line represents the median of the multiple values for each nx*nfacet levels."}
G21 %>% 
  ggplot(aes(x=nx*nfacet, y = value)) +
  geom_point() + stat_summary(fun=median, geom="line", aes(group=1), color = "blue") + xlab("nx*nfacet") + ylab("raw wpd")

@@

We need a transformation on $raw_{wpd}$ which will make it independent of the values of $nx*nfacet$. Two approaches have been employed for that purpose, the first one involves fitting a model and the latter involves a permutation method to make the distribution of the transformed $wpd$ similar across different $nx$ and $nfacet$.

@@

A log-linear model is fitted to see how the values of $wpd_{raw}$ changes with the values of $nx$ and $nfacet$. The model is of the form $$y = a+b*log(x) +e$$, where $y = median(wpd_{raw})$ and $x = nx*nfacet$. $wpd_{lm}$ is a transformation on $wpd_{raw}$ which should be designed to remove the effect of $nx*nfacet$ on $wpd_{raw}$ and thus is defined as follows:
$wpd_{lm} = wpd_{raw} - \hat a - \hat b*log(nx*nfacet)$

```{r wpd-lm-horizontal,echo=FALSE, fig.caption = "$wpd_{lm}$ is plotted again log(nx*nfacet) and this transformation leads to median($wpd_{lm}$) to having almost no relationship with log(nx*nfacet)"}

@@

#summary(fit_lm2)

@@

In the earlier approach, $wpd_{raw} \in R$ was assumed, whereas,  $0 \leq wpd_{raw}\leq 1$ since it is a Jensen-Shannon Distance (JSD). Further, we know JSD follows a Chi-square distribution, which is a special case of Gamma distribution and hence belongs to exponential family of distributions. Therefore, we can fit a generalized linear model instead of a linear model to allow the response variable to follow a Gamma distribution. The inverse link is used when we know that the mean response to be bounded, which is applicable in our case since JSD will lie between between 0 and 1.

We fit the Gamma generalized linear model with inverse link which is of the form:  
$$y = 1/(a+b*log(x)) + e$$, where $y = median(wpd_{raw})$, $x = nx*nfacet$. Again, $wpd_{glm}$ which is a transformation on $wpd_{raw}$ and is designed to remove the effect of $nx*nfacet$ on $wpd_{raw}$ are chosen as the residuals and defined as
$wpd_{glm} = wpd_{raw} - 1/(\hat a + \hat b*log(nx*nfacet))$.

<!-- Please note: -->

<!-- - subtracting the intercept brings the location of the transformed variable to 0 for either case -->
<!-- - heterogeneity is higher for this case after normalization as compared to the earlier one. -->

@@

@@

G21_permutation %>% 
    ggplot() +
  geom_density(aes(x = wpd_permutation), 
                   fill = "blue") +
  facet_grid(nx~nfacet,
             labeller = "label_both") +
  theme(legend.position = "bottom") 
```

@@

 - Making the range of both the variables same by using min-max scaling method. In practice, however, we would only have one value of $wpd_{raw}$ which we need to transform using the modeling approach. Hence, min-max scaling approach could not be used here.

 - Standardizing the variables and expressing scores at standard deviation units. Again in practice, however, we would only have one value of $wpd_{raw}$ which we need to transform using the modeling approach. Hence, standardizing scores could not be used here as we do not have the mean and standard deviation of a series while using transformation using modeling.

- Make the location and scale of both the approaches similar so that they could be compared. Please note that the range of values could be different in this case, however location and scale are brought to same levels.)

The measure $wpd_{glm}$ seems to roughly follow a normal distribution except in the tails as could be seen in Figure \ref{fig:hist-qq} and the very method of permutation approach ensures that $wpd_permutation$ is also normally distributed. Further, they are brought to the similar scale and location and hence could be compared.

```{r hist-qq, fig.cap = "In panel a, the histogram of $wpd_glm_trans$ is plotted. In parl b, the QQ plot is shown with the theoretical quantiles on the x-axis and $wpd_glm_trans$ quantiles on the y-axis. The distribution looks symmetric and looks like normal except in the tails."}
hist <- ggplot(G21_glm) +
  geom_histogram(aes(x = wpd_glm_trans)) +
  ggtitle("a")

p <- ggplot(G21_glm, aes(sample = wpd_glm_trans))
qqplot <-  p + geom_qq() + geom_qq_line() + coord_fixed(ratio = 4/5) +
  ggtitle("b")

 # G21_glm$wpd_glm_trans %>% 
 #   range()
hist + qqplot
```

>>>>

---
title: "Compare permutation and scalar transformation approaches on simulated and real data"
author: "Sayani Gupta"
output:
  bookdown::pdf_book:
    #base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      eval = TRUE,
                      warning = FALSE,
                      message = FALSE)
knitr::read_chunk("example_transformations.R")
library(tidyverse)
library(readr)
library(ggplot2)
library(here)
```

# Simulated data

## Data generation

@@

@@

 - Plot the values of raw wpd against nx*nfacet to see the rough relationship (\ref{fig:quadratic}).

```{r quadratic, fig.cap = "The raw wpd is plotted against nx*nfacet and the blue line represents the median of the multiple values for each nx*nfacet levels."}
G21 %>% 
  ggplot(aes(x=nx*nfacet, y = value)) +
  geom_point() + stat_summary(fun=median, geom="line", aes(group=1), color = "blue") + xlab("nx*nfacet") + ylab("raw wpd")

@@

@@

A log-linear model is fitted to see how the values of $wpd_{raw}$ changes with the values of $nx$ and $nfacet$. The model is of the form $$y = a+b*log(x) +e$$, where $y = median(wpd_{raw})$ and $x = nx*nfacet$. $wpd_{lm}$ is a transformation on $wpd_{raw}$ which should be designed to remove the effect of $nx*nfacet$ on $wpd_{raw}$ and thus is defined as follows:
$wpd_{lm} = wpd_{raw} - a - b*log(nx*nfacet)$

```{r wpd-lm-horizontal,echo=FALSE, fig.caption = "$wpd_{lm}$ is platted again log(nx*nfacet) and this transformation leads to median($wpd_{lm}$) to having almost no relationship with log(nx*nfacet)"}

@@

summary(fit_lm2)

@@

In the earlier approach, $wpd_{raw} \in R$, whereas,  $0 \leq wpd_{raw}\leq 1$ since it is a JS distance. Also, JSD follows a Chi-square distribution, which could be considered as a Gamma distribution. Hence, to restrict the range of the measure between 0 and 1, and deviate from the normality assumption of the response variable in a OLS approach, we fit a generalized linear model (GLM) with the error distributed as a gamma distribution and link function as "inverse". Hence, the model is of the form  $$1/y = a+b*log(x) +e$$, where $y = median(wpd_{raw})$ and $x = nx*nfacet$. Again, $wpd_{glm}$ is a transformation on $wpd_{raw}$ which should be designed to remove the effect of $nx*nfacet$ on $wpd_{raw}$ and thus is defined as follows:
$wpd_{glm} = 1/wpd_{raw} - a - b*log(nx*nfacet)$.

Please note:

- subtracting the intercept brings the location of the transformed variable to 0 for either case
- heterogeneity is higher for this case after normalization as compared to the earlier one.

@@

The measure $wpd_{glm}$ seems to follow a normal distribution.

```{r}
ggplot(G21_glm) +
  geom_histogram(aes(x = wpd_glm))

p <- ggplot(G21_glm, aes(sample = wpd_glm))
p + stat_qq() + stat_qq_line()

 G21_glm$wpd_glm_trans %>% 
   range()

@@

G21_permutation %>% 
    ggplot() +
  geom_density(aes(x = perm_trans), 
                   fill = "blue") +
  facet_grid(nx~nfacet,
             labeller = "label_both") +
  theme(legend.position = "bottom") 
```

@@

 - Converting each scale to have the same lower and upper levels
 - Standardizing the variables and expressing scores at standard deviation units

- making the range of both the variables same (additional advantage that we can have a bound for the normalised measure which could be compared across normalizing approaches and data sets.)
